Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-29 numerical.tex
updating

%  In particular, we focus
% on results most relevant to our eventual application to protoplanetary disks.

% The reader is referred
% to the above works for a detailed look at different VSI modes. Here,
% we present selected results most relevant to our eventual application
% to protoplanetary disks.

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated from 
Eq. \ref{simple_growth}. We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
Appendix \ref{max_growth1}. The match in $\nu$ worsens for 
decreasing $\khat$ due to the neglect of global radial structure in
the analysis. Nevertheless, given the number of approximations made
in our analysis, the agreement is satisfactory.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_eigen_imag_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/compare_eigen_real_iso}
  \caption{Growth rate (top) and real frequency (bottom) of the
    fundamental VSI mode in the fiducial disk model with $\beta =
    10^{-3}$. 
    % a disk with $(\gamma,
    % \Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The solid (dotted) line is obtained numerically
    (analytically).  
    \label{iso_eigen_kx} 
  }
\end{figure}

@@

@@

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. The fundamental mode is
robust against the vertical boundary conditions considered as it
appears at the same position in all cases. Higher order modes along
the analytic lines have larger growth rates, but require larger
vertical domains, since they have increasing number of nodes  $z$.

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$, except with the possible emergence
of modes associated with boundary conditions. Thus, tracking the
fundamental mode is a more preferable assessment of stability as $\beta$
is increased from zero, rather than searching for the most unstable
mode. This is because the latter may correspond to modes that depend on boundary
conditions, e.g. surface modes whose growth rate varies with
$\zmax$ in a vertically isothermal disk.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
We now test the critical thermal relaxation timescale given by
Eq. \ref{iso_vsi_cond}. We iterate starting from eigenvalues for the fundamental
mode with $\beta=10^{-3}$ and slowly increase the thermal
relaxation timescale up to $\beta=10^3$. Fig. \ref{bcrit_compare1}
show the growth rates obtained this way, together with the critical  
thermal relaxation timescale $\beta_\mathrm{crit}=0.125$ for our
fiducial disk parameters.

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides
an accurate prediction of the upper limit to the thermal relaxation 
timescale for the fundamental mode. For smaller $\khat$, the VSI can
operate slightly beyond $\beta_\mathrm{crit}$ (recall that
$\beta_\mathrm{crit}$ was derived for $\khat^2\gg 1$). For
$\khat\geq 50$, we find perturbations have large amplitudes near the
vertical boundaries, and 
complications to our numerical search of
eigenvalues arise from the emergence of modes associated with boundary
conditions. %We generally find boundary effects become more significant
%with increasing $\khat$.  
Nevertheless, growth rates rapidly decrease as $\beta$
approaches and increase beyond $\beta_\mathrm{crit}$.

Fig. \ref{bcrit_compare1} is consistent with numerical simulations
performed by \cite{nelson13} for which a thermal relaxation timescale
$\beta\gtrsim 0.6$ stabilized the VSI. This can be anticipated from our
estimate of the critical thermal timescale $\beta_\mathrm{crit}\simeq
0.1$ for our fiducial disk parameters (which is the same as that
adopted by \citeauthor{nelson13}). It is interesting to note that there is a
thermal timescale that maximizes the mode decay rate. Here,
it is $\beta\simeq1$---$2$ or 
about $0.1$---$0.3$ orbital periods. This is in fact consistent with 
Fig. 12 of \cite{nelson13} where the perturbed kinetic energy in their
numerical simulations were observed to decay most rapidly for a
thermal timescale of $\sim 0.1$ orbits.

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{VSI growth rates in the fiducial disk
     model as a function of the thermal relaxation timescale
     $\beta$. These are obtained by following the eigenvalues of the
     fundamental mode starting from $\beta=10^{-3}$ and increasing
     $\beta$. The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

 We further test the dependence of the critical thermal relaxation timescale 
 $\beta_\mathrm{crit}$ on disk parameters (Eq. \ref{iso_vsi_cond}).
 We use the fiducial setup as a  reference point and vary
 $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$ 
 separately.

@@

 Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 our finding that the fundamental mode becomes dominant as one
 increases $\beta$, and that it is not sensitive to boundary
 conditions, shows that $\beta_\mathrm{crit}$ as given  by
 Eq. \ref{iso_vsi_cond} provides a practical measure for the thermal
 relaxation timescale below which  VSI will operate in a vertically
 isothermal disk.

@@

Fig. \ref{compare_eigenvz_vnoniso_kx100}

>>>>

% Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
% for the fundamental mode and that calculated from 
% Eq. \ref{simple_growth}. We find good agreement in the
% wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
% $\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
% are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
% Appendix \ref{max_growth1}. The match in $\nu$ worsens for 
% decreasing $\khat$ due to the neglect of global radial structure in
% the analysis. Nevertheless, given the number of approximations made
% in our analysis, the agreement is satisfactory.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
%   0cm]{figures/compare_eigen_imag_iso} 
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   1cm]{figures/compare_eigen_real_iso}
%   \caption{Growth rate (top) and real frequency (bottom) of the
%     fundamental VSI mode in the fiducial disk model with $\beta =
%     10^{-3}$. 
%     % a disk with $(\gamma,
%     % \Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
%     % $\beta=10^{-3}$. 
%     The solid (dotted) line is obtained numerically
%     (analytically).  
%     \label{iso_eigen_kx} 
%   }
% \end{figure}

@@

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. The fundamental mode is
robust against the vertical boundary conditions considered as it
appears at the same position in all cases.

@@

Higher order modes along
the analytic lines have larger growth rates, but require larger
vertical domains, since they have increasing number of nodes  $z$.

@@

% Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
% show that the fundamental mode is not sensitive to boundary
% conditions. The fundamental mode also becomes dominant as $\beta$
% approaches $\beta_\mathrm{crit}$, except with the possible emergence
% of modes associated with boundary conditions. Thus, tracking the
% fundamental mode is a more preferable assessment of stability as $\beta$
% is increased from zero, rather than searching for the most unstable
% mode. This is because the latter may correspond to modes that depend on boundary
% conditions, e.g. surface modes whose growth rate varies with
% $\zmax$ in a vertically isothermal disk.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
In Fig. \ref{bcrit_compare1} we plot the maximum VSI growth rates
found in the fiducial disk model as a function of the thermal
relaxation timescale $\beta$ for several values of $\khat$. 
We also plot the critical timescale $\beta_\mathrm{crit}=0.125$ for
the fiducial disk parameters.

Note that the curves are not neccessarily smooth because the most unstable
mode may not always correspond to the same type of disturbance. For example,
as seen above, for $\khat \leq 10$ higher order body modes are
dominant as $\beta\to 0$, but the fundamental corrugation
mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
$\khat=5,\,10$ we find $\beta_\mathrm{crit}$ accurately predicts the
thermal timescal beyond which the VSI is suppressed.

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ we find for $\beta\gtrsim 0.1$ the dominant
disturbance correspond to the leftmost eigenvalues
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which depend on $\zmax$. For $\khat=50,\,100$ we find surface modes
dominate.

Our result is consistent with numerical simulations performed by
\cite{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
rates for $\beta\gtrsim 0.6$ are small. One might consider 
$\khat > 100$ to increase the maximum growth rate (note that the most unstable
mode for large $\khat$ are typically surface modes), but in a realistic disk
very small scale disturbances will likely be subject to viscous decay
(\S\ref{application}). Numerical simulations are also unlikely to
resolve such small scales.

As pointed out by \cite{barker15} surface modes
only appear in vertically isothermal disks when vertical boundaries
are imposed, so they are strongly dependent on $\zmax$.  Thus, seeking
the most unstable mode in a vertically isothermal disk may not yield
robust results because such modes, which are often surface modes, may
depend on boundary conditions.

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare_iso_maxrate} 
   \caption{Maximum VSI growth rates in the fiducial disk
     model as a function of the thermal relaxation timescale
     $\beta$. The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

Despite this caveat, and that surface modes are not accounted in our
analysis,  we see from Fig. \ref{bcrit_compare1} that 
$\beta_\mathrm{crit}$ nevertheless provides a reasonable estimate for
an upper limit to the thermal timescale that allows the VSI. We thus
test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately. We consider the fundamental
corrugation mode for which $\beta_\crit$ is the actual critical
thermal timescale.%  We use $\khat=10$ to avoid complications to our
% eigenvalue search from surface modes.

@@

Having validated

 % Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 % our finding that the fundamental mode becomes dominant as one
 % increases $\beta$, and that it is not sensitive to boundary
 % conditions, shows that $\beta_\mathrm{crit}$ as given  by
 % Eq. \ref{iso_vsi_cond} provides a practical measure for the thermal
 % relaxation timescale below which  VSI will operate in a vertically
 % isothermal disk.

@@

Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
$0.1$. At such large $\khat$, we find with $\beta=10^{-3}$ the most
unstable mode is an asymmetric surface mode \citep[also seen in Fig. 7
of][]{barker15}. However, increasing the thermal timescale to
$\beta=10^{-2}$ an anti-symmetric surface mode becomes
dominant. Further increasing to $\beta=0.1$ surface modes are
suppressed and the dominant mode is the fundamental corrugation
mode. Notice as $\beta$ is increased the perturbation amplitude near
the vertical boundaries become negligible. This is not surprsing since
the stabilizing effect of buoyancy is largest at the disk surface.

### 2015-03-29 paper.tex
updating

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that in a stably stratified disk,
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}
